A Unified Treatment of Tractability of Approximation Problems Defined on Hilbert Spaces

### 2023-07-19 Older_versions/GeneralizedTractability_230607.tex
Updates from Overleaf

The information complexity of the problem $\SOL_d$ is given by the function $\comp:(0,\infty) \times \naturals \to \natzero$, where $\comp(\varepsilon,d)$ is the number of linear functionals required by the best \peter{admissible} algorithm to satisfy the error criterion \eqref{eq:errcrit}.
By definition, the complexity is non-decreasing  as $\varepsilon$ tends to zero.  We also expect it to be non-decreasing as $d$ tends to infinity, since $d$ is typically the number of variables for the functions in $\cf_d$.

@@

In this article, we investigate a \emph{generalized tractability function}, denoted by $T$:
\begin{equation} \label{eq:Tspec}
    T :(0,\infty) \times \mathbb{N} \times [0,\infty)^s \rightarrow (0,\infty).
\end{equation}
The conditions we require $T$ to satisfy are described in  Section \ref{sec:not}, however, the basic idea is that we define our approximation problem to be tractable if $\comp(\varepsilon, d) \le  C_{\vp}\, T(\varepsilon^{-1},d,\vp)$ for some constant, $C_{\vp}$, depending only on the parameter $\vp$. \kachi{The parameter $\vp$ is a $s$-dimensional vector, where $\vp \geq 0$ implies that each component of $\vp$ is nonnegative.} In practice, we expect $T(\varepsilon^{-1},d,\vp)$ to increase sub-exponentially  with increasing $\varepsilon^{-1}$ and/or $d$, however, most of our theorems do not require such an assumption.

@@

\peter{TODO: Maybe add a short paragraph on the notation used in the paper and on its structure? In particular, we should introduce
an ordering of the parameters $\vp$.}[\kachi{KACHI: I will take care of this}]

\kachi{[KACHI: I will write about notations used in this paper. Is there are need to change notations like $[1,\infty)$ to $\naturals$ and $[0,\infty)$ to $\naturals_0$]}

\kachi{We denote the set of positive integers by $\naturals$ and and the set of non-negative integers by $\naturals_0$.}
In Section \ref{sec:spt}, we give the definitions of the considered tractability notions, in particular (strong) tractability. We consider strong tractability when the information complexity is independent of the dimension, $d$. We then turn to tractability  when the information complexity \peter{may depend on both} $d$ and $\varepsilon^{-1}$. In Section \ref{sec:examples} we provide examples of \peter{tractability functions} for both the algebraic and exponential cases. We also consider other notions of tractability such as quasi-polynomial tractability, and a generalized notion of weak tractability. In the case of generalized weak tractability, we give the necessary and sufficient conditions on the notion of exponential tractability in Section \ref{sec:subh} \peter{[I think I do not quite understand what we mean by the last sentence.]}.

@@

\begin{proof}
\textbf{Sufficient condition:}\newline
We make the first part of the argument in some generality so that it can be reused in the proof of Theorem \ref{thm_main_tract2} for tractability.  Fix any $\varepsilon > 0$ and any $d \in \naturals$.
\kachi{[KACHI: It seems the notation of the dimension has change from $d$ to $\delta$. I will work on this.]}
Since the $\lambda_{i,d}$ are non-increasing in $i$, it follows that the $T(\lambda_{i,d}^{-1},d,\vp)$ are non-decreasing in $i$ for any $\thed\in\naturals$. In particular, for  any $N\in\naturals$ with $n\ge N$, we have
\begin{align*}
    \lambda_{n+1,d} \le \lambda_{n,d} \le \cdots \le \lambda_{N,d}
    & \implies \frac{1}{T(\lambda_{n+1, d}^{-1},\thed,\vp)} \le \frac{1}{T(\lambda_{n, d}^{-1},\thed,\vp)} \le \cdots \le \frac{1}{T(\lambda_{N, d}^{-1},\thed,\vp)} \\
     \implies \frac{1}{T(\lambda_{n+1, d}^{-1},\thed,\vp) }
    \le &\frac{1}{n-N+1} \sum_{i=N}^n  \frac{1}{T(\lambda_{i, d}^{-1},\thed,\vp) }
    \le \frac{1}{n-N+1} \sum_{i=N}^\infty  \frac{1}{T(\lambda_{i, d}^{-1},\thed,\vp)}.
\end{align*}
Thus, we can conclude from the previous line that
\begin{multline} \label{eq:implicate_a}
    n - N +1 \ge T(\varepsilon^{-1},\thed,\vp) \sum_{i=N}^\infty \frac{1}{T(\lambda_{i, d}^{-1},\thed,\vp)} \\
   \implies   \frac{1}{T(\lambda_{n+1, d}^{-1},\thed,\vp)} \le
   \frac{1}{n-N+1} \sum_{i=N}^\infty \frac{1}{T(\lambda_{i, d}^{-1},\thed,\vp) } \le \frac{1}{T(\varepsilon^{-1},\thed,\vp)} \\ \forall \thed, n, N \in \naturals.
\end{multline}
We note that the left side of the implication implies $n \ge N$ automatically since $n$ is an integer  $n - N +1$ is assumed to be positive.

Moreover, given that $T$ is increasing in its arguments, we have the following equivalent expression for the complexity via \eqref{eq:comp_def}:
\begin{align} %\label{eq:comp_equiv}
	\nonumber
	\comp(\varepsilon, d) & = \min \{n \in \natzero : \lambda_{n+1, d} \le \varepsilon\} \\
	\nonumber
	& = \min \biggl\{n \in \natzero : \frac{1}{T(\lambda_{n+1,d}^{-1},\thed,\vp)}\le \frac{1}{T(\varepsilon^{-1},\thed,\vp) }\biggr\} \qquad  \forall \thed\in \naturals\\
	\nonumber
	& \le  \min \biggl\{n \in \natzero : n \ge N - 1 + T(\varepsilon^{-1},\thed,\vp) \sum_{i=N}^\infty \frac{1}{T(\lambda_{i, d}^{-1},\thed,\vp)} \biggr\} \;  \forall  \thed, N \in \naturals, \quad \text{by \eqref{eq:implicate_a}} \\ 
	& \le  T(\varepsilon^{-1},\thed,\vp) \Biggl [\frac{N}{T(\varepsilon^{-1},\thed,\vp)}  +  \sum_{i=N}^\infty \frac{1}{T(\lambda_{i, d}^{-1},\thed,\vp)} \Biggr] \quad \forall  \thed, N \in \naturals. \label{eq:compUP}
\end{align}

Now we take this upper bound further, specializing to the case of $\thed=1$ and $N=L_{\vp}$:
\begin{align*}
       \comp(\varepsilon,d)
       & \le T(\varepsilon^{-1},1, \vp) \Biggl [ \frac{L_{\vp}}{T(\varepsilon^{-1},1, \vp)}  + \underbrace{\sum_{i=L_{\vp}}^\infty \frac{1}{T(\lambda_{i, d}^{-1},1,\vp)}}_{\le S_{\vp} <\infty\ \text{by \eqref{eq:strong_tractiff3}} }
        \Biggr] \\
       & \le T(\varepsilon^{-1},1,\vp) \underbrace{\left[\frac{L_{\vp}}{T(0,1, \vp)} + S_{\vp} \right]}_{=:C_{\vp}}
       \qquad \text{by \eqref{eq:Tnondecreasing}}\\
       & =  C_{\vp}T(\varepsilon^{-1},1,\vp).
%       & \le \underbrace{2 C_{1,\vp}}_{\text{since } T(\varepsilon^{-1},1,\vp) \ge 1} T(\varepsilon^{-1},1,\vp)
%       = \underbrace{C_{\vp}}_{:=2 C_{1,\vp}}T(\varepsilon^{-1},1,\vp).
\end{align*}
This means that we have strong $T$-tractability via \eqref{eq:strong_tract_def}, and verifies the sufficiency of \eqref{eq:strong_tractiff3}.

@@

\begin{example}[Weak Tractability]
	Suppose that
	\[
	T(\varepsilon^{-1},d,s,t) = \max(1,\varepsilon^{-1})^s + d^t, \quad \forall \varepsilon,s,t > 0, \, d \in \naturals, \qquad h(x) = \exp(x) \quad \forall x \ge 0.
	\]
	Then Theorem \ref{thm:subhT} implies that
	\[\lim_{\varepsilon^{-1}+d\rightarrow \infty} \frac{\log( \comp(\varepsilon,d))}{\varepsilon^{-s}+d^t} = 0
	\iff
	 \sup_{d\in\naturals}  \, \sum_{i=1}^\infty \exp(-c\, (\min(1,\lambda_{i,d})^{s} + d^t )) <\infty \qquad \forall c > 0.
	\]
	The expression on the left is the notion of $(s,t)$-weak tractability \peter{introduced in \cite{SW15}}, and the condition on the right was derived in \peter{\cite[Theorem 3.1]{WW17}}.  Weak tractability corresponds to $(s,t)=(1,1)$, and uniform weak tractability corresponds to the case when a problem is weakly tractable for all positive $s$ and $t$.
\end{example}

>>>>

The information complexity of the problem $\SOL_d$ is given by the function $\comp:(0,\infty) \times \naturals \to \natzero$, where $\comp(\varepsilon,d)$ is the number of linear functionals required by the best \peter{admissible} algorithm to satisfy the error criterion \eqref{eq:errcrit}.
By definition, the \kachi{information} complexity is non-decreasing  as $\varepsilon$ tends to zero.  We also expect it to be non-decreasing as $d$ tends to infinity, since $d$ is typically the number of variables for the functions in $\cf_d$.

@@

In this article, we investigate a \emph{generalized tractability function}, denoted by $T$:
\begin{equation} \label{eq:Tspec}
    T :(0,\infty) \times \mathbb{N} \times [0,\infty)^s \rightarrow (0,\infty).
\end{equation}
The conditions we require $T$ to satisfy are described in  Section \ref{sec:not}, however, the basic idea is that we define our approximation problem to be tractable if $\comp(\varepsilon, d) \le  C_{\vp}\, T(\varepsilon^{-1},d,\vp)$ for some constant, $C_{\vp}$, depending only on the parameter $\vp$. \kachi{The parameter $\vp$ is a $s$-dimensional vector, for $s \geq 1$, where $\vp \geq 0$ implies that each component of $\vp$ is non-negative.} In practice, we expect $T(\varepsilon^{-1},d,\vp)$ to increase sub-exponentially  with increasing $\varepsilon^{-1}$ and/or $d$, however, most of our theorems do not require such an assumption.

@@

\kachi{In Table \ref{tab:commonT} below we give common forms of the tractability function, $T$ considered in this paper.}

%\peter{TODO: Maybe add a short paragraph on the notation used in the paper and on its structure? In particular, we should introducean ordering of the parameters $\vp$.}[\kachi{KACHI: I will take care of this}]

%\kachi{[KACHI: I will write about notations used in this paper. Is there are need to change notations like $[1,\infty)$ to $\naturals$ and $[0,\infty)$ to $\naturals_0$]}

\kachi{We denote the set of positive integers by $\naturals$ and and the set of non-negative integers by $\naturals_0$. The remainder of this paper is organized as follows.}
In Section \ref{sec:spt}, we give the definitions of the considered tractability notions, in particular (strong) tractability. We consider strong tractability when the information complexity is independent of the dimension, $d$. We then turn to tractability  when the information complexity \peter{may depend on both} $d$ and $\varepsilon^{-1}$. In Section \ref{sec:examples} we provide examples of \peter{tractability functions} for both the algebraic and exponential cases. We also consider other notions of tractability such as quasi-polynomial tractability, and a generalized notion of weak tractability. In the case of generalized weak tractability, we give the necessary and sufficient conditions on the notion of exponential tractability in Section \ref{sec:subh} \peter{[I think I do not quite understand what we mean by the last sentence.]}.

@@

\begin{proof}
\textbf{Sufficient condition:}\newline
We make the first part of the argument in some generality so that it can be reused in the proof of Theorem \ref{thm_main_tract2} for tractability.  Fix any $\varepsilon > 0$ and any $d \in \naturals$.
\kachi{[KACHI: It seems the notation of the dimension has change from $d$ to $\delta$. I will work on this.]}
Since the $\lambda_{i,d}$ are non-increasing in $i$, it follows that the $T(\lambda_{i,d}^{-1},d,\vp)$ are non-decreasing in $i$ for any $d\in\naturals$. In particular, for  any $N\in\naturals$ with $n\ge N$, we have
\begin{align*}
    \lambda_{n+1,d} \le \lambda_{n,d} \le \cdots \le \lambda_{N,d}
    & \implies \frac{1}{T(\lambda_{n+1, d}^{-1},d,\vp)} \le \frac{1}{T(\lambda_{n, d}^{-1},d,\vp)} \le \cdots \le \frac{1}{T(\lambda_{N, d}^{-1},d,\vp)} \\
     \implies \frac{1}{T(\lambda_{n+1, d}^{-1},d,\vp) }
    \le &\frac{1}{n-N+1} \sum_{i=N}^n  \frac{1}{T(\lambda_{i, d}^{-1},d,\vp) }
    \le \frac{1}{n-N+1} \sum_{i=N}^\infty  \frac{1}{T(\lambda_{i, d}^{-1},d,\vp)}.
\end{align*}
Thus, we can conclude from the previous line that
\begin{multline} \label{eq:implicate_a}
    n - N +1 \ge T(\varepsilon^{-1},d,\vp) \sum_{i=N}^\infty \frac{1}{T(\lambda_{i, d}^{-1},d,\vp)} \\
   \implies   \frac{1}{T(\lambda_{n+1, d}^{-1},d,\vp)} \le
   \frac{1}{n-N+1} \sum_{i=N}^\infty \frac{1}{T(\lambda_{i, d}^{-1},d,\vp) } \le \frac{1}{T(\varepsilon^{-1},d,\vp)} \\ \forall d, n, N \in \naturals.
\end{multline}
We note that the left side of the implication implies $n \ge N$ automatically since $n$ is an integer  $n - N +1$ is assumed to be positive.

Moreover, given that $T$ is increasing in its arguments, we have the following equivalent expression for the complexity via \eqref{eq:comp_def}:
\begin{align} %\label{eq:comp_equiv}
	\nonumber
	\comp(\varepsilon, d) & = \min \{n \in \natzero : \lambda_{n+1, d} \le \varepsilon\} \\
	\nonumber
	& = \min \biggl\{n \in \natzero : \frac{1}{T(\lambda_{n+1,d}^{-1},d,\vp)}\le \frac{1}{T(\varepsilon^{-1},d,\vp) }\biggr\} \qquad  \forall d\in \naturals\\
	\nonumber
	& \le  \min \biggl\{n \in \natzero : n \ge N - 1 + T(\varepsilon^{-1},d,\vp) \sum_{i=N}^\infty \frac{1}{T(\lambda_{i, d}^{-1},d,\vp)} \biggr\} \;  \forall  d, N \in \naturals, \quad \text{by \eqref{eq:implicate_a}} \\ 
	& \le  T(\varepsilon^{-1},d,\vp) \Biggl [\frac{N}{T(\varepsilon^{-1},d,\vp)}  +  \sum_{i=N}^\infty \frac{1}{T(\lambda_{i, d}^{-1},d,\vp)} \Biggr] \quad \forall  d, N \in \naturals. \label{eq:compUP}
\end{align}

Now we take this upper bound further, specializing to the case of $d=1$ and $N=L_{\vp}$:
\begin{align*}
       \comp(\varepsilon,d)
       & \le T(\varepsilon^{-1},1, \vp) \Biggl [ \frac{L_{\vp}}{T(\varepsilon^{-1},1, \vp)}  + \underbrace{\sum_{i=L_{\vp}}^\infty \frac{1}{T(\lambda_{i, d}^{-1},1,\vp)}}_{\le S_{\vp} <\infty\ \text{by \eqref{eq:strong_tractiff3}} }
        \Biggr] \\
       & \le T(\varepsilon^{-1},1,\vp) \underbrace{\left[\frac{L_{\vp}}{T(0,1, \vp)} + S_{\vp} \right]}_{=:C_{\vp}}
       \qquad \text{by \eqref{eq:Tnondecreasing}}\\
       & =  C_{\vp}T(\varepsilon^{-1},1,\vp).
%       & \le \underbrace{2 C_{1,\vp}}_{\text{since } T(\varepsilon^{-1},1,\vp) \ge 1} T(\varepsilon^{-1},1,\vp)
%       = \underbrace{C_{\vp}}_{:=2 C_{1,\vp}}T(\varepsilon^{-1},1,\vp).
\end{align*}
This means that we have strong $T$-tractability via \eqref{eq:strong_tract_def}, and verifies the sufficiency of \eqref{eq:strong_tractiff3}.

@@

\begin{example}[Weak Tractability]
	Suppose that
	\[
	T(\varepsilon^{-1},d,\kachi{(s,t)}) = \max(1,\varepsilon^{-1})^s + d^t, \quad \forall \varepsilon,s,t > 0, \, d \in \naturals, \qquad h(x) = \exp(x) \quad \forall x \ge 0.
	\]
	Then Theorem \ref{thm:subhT} implies that
	\[\lim_{\varepsilon^{-1}+d\rightarrow \infty} \frac{\log( \comp(\varepsilon,d))}{\varepsilon^{-s}+d^t} = 0
	\iff
	 \sup_{d\in\naturals}  \, \sum_{i=1}^\infty \exp(-c\, (\min(1,\lambda_{i,d})^{s} + d^t )) <\infty \qquad \forall c > 0.
	\]
	The expression on the left is the notion of $(s,t)$-weak tractability \peter{introduced in \cite{SW15}}, and the condition on the right was derived in \peter{\cite[Theorem 3.1]{WW17}}.  Weak tractability corresponds to $(s,t)=(1,1)$, and uniform weak tractability corresponds to the case when a problem is weakly tractable for all positive $s$ and $t$.
\end{example}
